---
title: "Practical Exercise: Multiple Regression and Model Evaluation"
subtitle: "Homework Deadlines and Student Performance"
author: "Tatjana Kecojevic"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
    tbl-colwidths: [50, 110]
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

In this practical, you will extend your analysis of the homework dataset by fitting and interpreting a **multiple regression model**.

Last week, we focused on simple relationships between two variables.

This week, we move one step further and ask:

> **What happens when we consider several factors at the same time?**

Our goal is to:

-   build a multiple regression model
-   interpret its results
-   evaluate whether the model is useful

# Step 1: Load and Prepare the Data

Before we begin, we need to make sure the data is in the correct format.

```{r}
# Load the dataset
hw <- read.csv("https://raw.githubusercontent.com/TanjaKec/mydata/master/HW_R.csv")

hw$HW_minutes <- as.numeric(hw$HW_minutes)
hw$GPA <- as.numeric(hw$GPA)
hw$ACT <- as.numeric(hw$ACT)
hw$Q3_deadline_stress <- as.numeric(hw$Q3_deadline_stress)
hw$Midnight_deadline <- as.factor(hw$Midnight_deadline)

# Inspect structure
str(hw)
```

# Step 2: Explore the Data

Before fitting any model, it is always a good idea to look at the data first. This helps us build intuition about the relationships we might expect to see.

```{r}
library(GGally)

vars <- hw[, c("Grade_course", "HW_minutes", "GPA", "ACT", "Q3_deadline_stress")]
ggpairs(vars)
```

Take a moment to scan the plots. You are not looking for exact answers yet, just general patterns.

Questions:

1.  Which variables seem related to `Grade_course`? \vspace{2cm}
2.  Are any predictors related to each other? \vspace{2cm}
3.  Are the relationships strong or weak? \vspace{2cm}

# Step 3: Fit the Model

Now we move from exploration to modelling. Instead of looking at variables one at a time, we include them all together in a single model.

```{r}
model_multi <- lm(
  Grade_course ~ HW_minutes + GPA + ACT + Midnight_deadline + Q3_deadline_stress,
  data = hw
)

summary(model_multi)
```

This allows us to ask a more realistic question:

> **How does each variable relate to course performance once we account for the others?**

Step 4: Coefficients

The first part of the output tells us about the individual relationships. Each coefficient represents the effect of a variable holding the others constant.

Questions:

1.  What is the estimated effect of HW_minutes? \vspace{1.5cm}
2.  What does the coefficient of GPA mean? \vspace{1.5cm}
3.  Do the signs of the coefficients make sense? \vspace{2cm}

Take a moment to compare these results to what you saw in the plots.

# Step 5: Model Fit

Next, we step back and look at the model as a whole. We ask:

> **How much of the variation in course performance does this model actually explain?**

Questions:

1.  What is $R^2$? \vspace{1cm}
2.  What is $R^2_{\text{adj}}$? \vspace{1cm}
3.  Why is adjusted $R^2$ lower? \vspace{2cm}

This is an important idea:

-   $R^2$ tells us how well the model fits
-   $R^2_{\text{adj}}$ tells us how much of that fit we can trust

# Step 6: The F-test

We now move from description to formal testing. The F-test asks:

> **Does this model, taken as a whole, explain anything?**

Questions:

1.  What is the F-statistic? \vspace{1cm}
2.  What is the p-value? \vspace{1cm}
3.  Do we reject $H_0$? \vspace{2cm}

Interpret your answer:

Does the model provide evidence of a relationship with Grade_course? \vspace{2cm}

# Step 7: Individual t-tests

Even if the overall model is weak, we can still look at individual variables.

Each t-test asks:

> **Does this variable contribute to the model, once the others are included?**

Questions:

1.  Which variables are significant? \vspace{2cm}
2.  Do any predictors clearly stand out? \vspace{2cm}
3.  How do these results compare to what you saw earlier? \vspace{2cm}

# Final Reflection

Now step back and look at the full picture.

What do all of these results tell us? \vspace{3cm}

You may find that:

-   relationships exist, but are weak
-   no single variable dominates
-   the model explains only a small amount of variation

This is not unusual. This is not a failure of regression, but a reminder that simple models cannot always capture complex reality.

# Looking Ahead

This raises an important question:

Is the issue the data, or the model?

In the next session, we will build on this by exploring:

-   multicollinearity (when predictors overlap)
-   categorical variables and dummy coding
-   non-linear relationships
-   interactions between variables

The goal is to move from a simple model to a more realistic one.
