---
title: "Building and Refining Multiple Regression Models"
subtitle: "From Specification to Simplification: Developing Regression Models"
author: "Tatjana Kecojević"
format: 
  pdf:
    documentclass: article
    fontsize: 11pt
    tbl-colwidths: [50, 110]
execute:
  echo: true
  warning: false
  message: false
---

## 1. Introduction

In this practical, we move from understanding regression models to **building and refining them ourselves**.

Rather than reproducing the models from the lecture, you will take an active role as a **modeller**. You will:

- construct your own regression models  
- assess their fit  
- interpret results  
- simplify models where appropriate  

There is **no single correct answer**; what matters is your reasoning.

::: {.callout-important}
**Thinking like a modeller**

Model building is not mechanical. You must balance:   
- statistical significance  
- model fit ($R^2_{\text{adj}}$)  
- simplicity (parsimony)  

Your goal is to justify your decisions.
:::

::: {.callout-important}
**Important**

There are **no model answers** for this practical.

This is intentional. The aim is to help you develop the skills needed to:   
- make your own modelling decisions  
- justify your choices  
- interpret results independently  

Different students may arrive at different models, this is expected.

This practical is designed to help you prepare for your **assessment**, where you will be required to carry out a similar analysis on your own.
:::

---

## 2. Loading the data

```{r}
library(wooldridge)
library(tidyverse)

data("wage1")

glimpse(wage1)
```
We will model wage using a selection of explanatory variables.

## 3. Building your first model

**Choosing your variables**

In the lecture, we focused on a specific set of predictors (e.g. `educ`, `exper`, `tenure`, `female`, `married`, `nonwhite`).

In this practical, you are encouraged to go beyond these choices.

Explore the dataset and consider including additional variables, such as:     

- `numdep`, number of dependents
- `smsa`, urban location indicator
- regional indicators (`northcen`, `south`, `west`)
- occupation or industry variables (e.g. `trade`, `services`, `profocc`)\

::: {.callout-important}
**Important**

Do not simply replicate the model from the lecture.

You are encouraged to:    

- experiment with different combinations of variables
- compare alternative specifications
- justify your modelling choices

Different students may obtain different models; this is expected.
:::

### Task 1

Construct a regression model using:   

- at least three predictors, and
- at least one variable not used in the lecture model\

```{r, eval = FALSE}
model_1 <- lm(wage ~ ..., data = wage1)
summary(model_1)
```

**Questions**

1. What is $R^2$?
    \vspace{1cm}
2. What is $R^2_{\text{adj}}$?
    \vspace{1cm}
3. Which variables are statistically significant?
    \vspace{2cm}
    
::: {.callout-note}
**Tip**

Before building your model, use `glimpse()` or `summary()` to explore the dataset and identify variables that might plausibly affect wages.
:::
\vspace{0.15cm}
::: {.callout-warning}
**Challenge (optional)**

Try including a variable whose effect is not immediately obvious.

For example, you might explore:

- `smsa` (whether the individual lives in an urban area)

Questions to consider:

- Does it turn out to be statistically significant?  
- Does it change the results for other variables?  
- Does its effect make sense in context?  

Most importantly:

> Does this variable help explain the variability in **wage**?

Remember, your goal is to explain differences in the outcome variable ($Y$). A useful predictor should contribute to explaining why wages vary across individuals.

This helps you think critically about model specification.
:::
\vspace{.3cm}

## Final note

::: {.callout-note}
**Explore and enjoy the process**

Have fun with this.

Try different models, change variables, and see what happens. There is a lot to learn by experimenting and observing how your results change.  

Play with the data, explore different ideas, and don’t be afraid to make mistakes, that’s how you develop intuition.  

The goal here is not just to get an answer, but to **understand the modelling process**.  

**Hints to guide your exploration** 
    
- Start simple, then gradually add variables  
- Pay attention to how $R^2_{\text{adj}}$ changes as you modify your model  
- Watch how coefficients change when you add or remove variables  
- Think about whether the signs and magnitudes make sense  
- Consider whether a variable is statistically significant *and* meaningful  
- If two models perform similarly, prefer the simpler one  

\vspace{0.25cm}
**Learn, experiment, and enjoy it!**
:::